---
title: Prediction Bands
author: Darrin Speegle
date: '2018-04-07'
slug: prediction-bands
categories: []
tags:
  - regression 
  - simulation
  - R
---

##Introduction

In this post, we study a seemingly easy question; what is a 95% prediction interval in simple linear regression? We assume that our data comes from 
$$
y_i = 1 + 2x_i + \epsilon_i
$$
where $\epsilon_i$ are iid normal with mean 0 and standard deviation 3. 

##R Code for Prediction

We can build a linear model and prediction intervals using the following R code:
```{r}
set.seed(4172018)
x <- runif(30, 0, 10)
y <- 1 + 2 * x + rnorm(30, 0, 3)
mod <- lm(y~x)
predict(mod, newdata = data.frame(x = 5), interval = "pre")
```
We see that we have an estimate of 12 (true value: $1 + 2 \times 5$), a lower prediction bound of 6.3 and an upper of 17.8. What percentage of future observations will fall in this interval?
```{r}
sim_dat <- replicate(10000, {
  new_point <- 1 + 2 * 5 + rnorm(1, 0, 3)
  new_point > 6.313055 && new_point < 17.76562
})
mean(sim_dat)
```
We see that 9281 out of 10,000 future data points fall in this 95% prediction interval. Since we know the true way of generating data, we can compute the exact probability that future data will fall in this interval via:
```{r}
pnorm(17.76562 , 11, 3) - pnorm(6.313055, 11, 3)
```
What's going on? How does constructing a 95% prediction interval lead to an interval that collects 92.88% of future data?

## Sampling Distribution

To understand this better, let's estimate (via simulation) the sampling distribution of the probability that a prediction interval will contain future values.
```{r}
sampling_dist <- replicate(1000, {
  x <- runif(30, 0, 10)
  y <- 1 + 2 * x + rnorm(30, 0, 3)
  mod <- lm(y~x)
  pre <- predict(mod, newdata = data.frame(x = 5), interval = "pre")
  pnorm(pre[3], 11, 3) - pnorm(pre[2], 11, 3)
})
hist(sampling_dist)
min(sampling_dist) #Smallest percentage of future values in prediction interval
mean(sampling_dist) #Expected percentage of future values in prediction interval
mean(sampling_dist < 0.9) #percentage of times that prediction interval predicts less than 90% of future values
density(sampling_dist)$x[which.max(density(sampling_dist)$y)] #Most likely percentage of future values contained in prediction interval
```

The histogram of attained probabilities that a prediction interval will contain future data explains what is going on. The probabilities have *expected value* of 0.95, but the distribution is skew. The most likely outcome seems to be that the prediction interval contain about 97% of future data, but there is about a 8% chance that it will contain less than 90% of future data. The minimum amount of future data contained (in this sample size of 1000) is 77%! 

## Sample Size

I have heard people say that the benefit of having more samples is that the prediction intervals get narrower. That is true, but also true is that a 95% prediction interval is also much more likely to contain about 95% of future values! Let's see what happens when we have 300 and 3000 data points, rather than 30:

```{r, echo = FALSE}
sample_size <- 300
sampling_dist <- replicate(1000, {
  x <- runif(sample_size, 0, 10)
  y <- 1 + 2 * x + rnorm(sample_size, 0, 3)
  mod <- lm(y~x)
  pre <- predict(mod, newdata = data.frame(x = 5), interval = "pre")
  pnorm(pre[3], 11, 3) - pnorm(pre[2], 11, 3)
})
hist(sampling_dist, main = "Probability with n = 300", probability = TRUE)
```

```{r, echo = FALSE}
sample_size <- 3000
sampling_dist <- replicate(1000, {
  x <- runif(sample_size, 0, 10)
  y <- 1 + 2 * x + rnorm(sample_size, 0, 3)
  mod <- lm(y~x)
  pre <- predict(mod, newdata = data.frame(x = 5), interval = "pre")
  pnorm(pre[3], 11, 3) - pnorm(pre[2], 11, 3)
})
hist(sampling_dist, main = "Probability with n = 3000", probability = TRUE)
```



## Table of Standard Deviations

Finally, we estimate the standard deviation of the sampling distribution of the probability that a future value will be in a prediction interval for various sample sizes.

```{r, cache = TRUE}
sample_sizes <- c(10, 20, 50, 100, 200, 500, 1000, 2000, 5000)
sds <- sapply(sample_sizes, function(n) {
  sd(replicate(1000, {
    x <- runif(n, 0, 10)
    y <- 1 + 2 * x + rnorm(n, 0, 3)
    mod <- lm(y~x)
    pre <- predict(mod, newdata = data.frame(x = 5), interval = "pre")
    pnorm(pre[3], 11, 3) - pnorm(pre[2], 11, 3)
  }))
})
knitr::kable(data.frame(`Sample Size` = sample_sizes, `Standard Dev` = sds))

```

